Error Reporting Parsers: a Monad Transformer Approach

### 2013-05-19 article.tex
parse 0 and monad transformer article sections: continue working on English of explanation, and put more imports by the code that uses them

Analagous to \verb+mfilter+ is the \verb+check+ combinator, which is used to
validate a parse result:
\begin{verbatim}
check :: (Monad f, Alternative f) => (a -> Bool) -> f a -> f a
check p m =
    m >>= \x -> if p x then return x else empty
\end{verbatim}

@@

*Main> getParser parseNest "(()abc(def)()"
Just ("(()abc(def)()",[])
\end{verbatim}
In the rest of this article, we'll be reworking these basic parser combinators,
by improving the implementations and extending the result types,
all while taking advantage of monad transformers to keep it modular.

\section{Monad transformers}
Monad transformers \cite{liang} allow monads to be modularly combined to form 
new, more complicated monads; the combined monads support all the operations of
their constituents.  Doing just that -- combining multiple monads into a stack --
we can build semantically rich yet modular parsers.
We'll try to use the standard transformers/mtl \cite{mtl} libraries as much as possible, 
since they're easy to obtain, easy to use, and quite effective in practice, but
occasionally we'll have to make a slight modification or extensions.

First, note that the parser type we used earlier -- \verb+[t] -> Maybe ([t], a)+
-- can be built out of the StateT monad transformer applied to Maybe.  We can 
now define the new parser type as:
\begin{verbatim}
type Parser t a = StateT [t] Maybe a
\end{verbatim}

We can build the \verb+item+ parser using the combinators from MonadState --
token lists are the 'state' that \verb+item+ must inspect and modify:
\begin{verbatim}
{-# LANGUAGE   FlexibleContexts  #-}

import Control.Monad.State       (MonadState (..), StateT(..))

item :: (MonadState [t] m, Alternative m) => m t
item =
    get >>= \xs -> case xs of
                        (t:ts) -> put ts *> pure t;
                        []     -> empty;
\end{verbatim}
Note that we needed to use the FlexibleContexts language extension for the 
MonadState constraints in the type signatures.  This formulation of monad
transformers relies heavily on extensions to the type system, more of which 
we'll have to use for some of the later code in this article.

>>>>

Analagous to \verb+Control.Monad.mfilter+ is the \verb+check+ combinator, which is used to
validate a parse result:
\begin{verbatim}
check :: (Monad f, Alternative f) => (a -> Bool) -> f a -> f a
check p m =
    m >>= \x -> if p x then return x else empty
\end{verbatim}

@@

*Main> getParser parseNest "(()abc(def)()"
Just ("(()abc(def)()",[])
\end{verbatim}
In the rest of this article, we'll rework and augment these basic parser 
combinators by genericizing the implementations and extending the result types,
while taking advantage of monad transformers to keep them modular.

\section{Monad transformers}
Monad transformers \cite{liang} allow monads to be modularly combined to form 
new, combined monads which support operations of each of their constituents.  
There are several useful monads for parsing, and by applying monad transformers, 
we can build semantically rich parsers without sacrificing modularity.
We'll use the standard transformers/mtl \cite{mtl} libraries -- 
they're easy to get, easy to use, and very useful.

First, note that the parser type we used earlier -- \verb+[t] -> Maybe ([t], a)+
-- can be built out of the StateT monad transformer applied to Maybe.  We can 
now define the new parser type as:
\begin{verbatim}
import Control.Monad.State       (StateT(..))

type Parser t a = StateT [t] Maybe a
\end{verbatim}

The \verb+item+ parser is reimplemented using the combinators from MonadState --
token lists are the 'state' that \verb+item+ must inspect and modify:
\begin{verbatim}
{-# LANGUAGE   FlexibleContexts  #-}

import Control.Monad.State       (MonadState (..))

item :: (MonadState [t] m, Alternative m) => m t
item =
    get >>= \xs -> case xs of
                        (t:ts) -> put ts *> pure t;
                        []     -> empty;
\end{verbatim}
Note that the FlexibleContexts language extension is required by the 
MonadState constraint in the type signature.  This formulation of monad
transformers relies heavily on extensions to the type system, more of which 
we'll have to use for some of the later code in this article.
